---
title: | 
    | Randomization (Design-based) Inference 
    | for Experiments
date: "2024-08-20"
date-format: iso
author: Ryan T. Moore
format:
  beamer:
    fonttheme: serif
    include-in-header:
      - file: zzz_beamer_header.tex
    section-titles: true
    slide_level: 2
    toc: true
institute: 
  - American University
link-citations: true 
bibliography: ../admin/main.bib
---
  

```{r}
#| label: enviro
#| echo: false
#| message: false
#| warning: false

library(kableExtra)
library(tidyverse)
```

# An Exercise

***

\Huge

\begin{center}
A volunteer?
\end{center}

\pause 

\large

\vspace{3mm}

The task: select the 2 folders with messages \pause

\vspace{3mm}
>- What is our baseline expectation/model for this process?  
>    - ``No x-ray vision. No ESP. Effect of messages on choice $=0$.\text{''}
>- What is an alternative?
>    - ``Some way to detect messages. Message location $\to$ choice.\text{''}

\pause 

\vspace{3mm}

Select!

## Randomization Inference

The possible choices:
\begin{center}
$\blacksquare \blacksquare \square \square \square$ \quad $\square \blacksquare \square \blacksquare \square$\\
$\blacksquare \square \blacksquare \square \square$ \quad $\square \blacksquare \square \square \blacksquare$\\
$\blacksquare \square \square \blacksquare \square$ \quad \alert<2->{$\square \square \blacksquare \blacksquare \square$}\\
$\blacksquare \square \square \square \blacksquare$ \quad $\square \square \blacksquare \square \blacksquare$\\
$\square \blacksquare \blacksquare \square \square$ \quad $\square \square \square \blacksquare \blacksquare$\\
\end{center}

\pause 

\begin{itemize}
	\item You chose $\_\_\_$ and $\_\_\_$.  Let $X=$ number found. \pause 
	\item What was $P(X \geq 2 | \text{no ESP})$?  \pause $\frac{1}{10} = 0.1$ \pause 
	\item What was $P(X \geq 1 | \text{no ESP})$?  \pause $\frac{7}{10} = 0.7$ \pause 
	\item What is ``prob result at least this extreme, given model of no effect\text{''}?  \pause 
	\item Definition of $p$-value! \pause 
	\item Valid, exact, with no distributional assumption, no large $n$.  \pause
	\item {\it Randomization} creates dist'n of possible numbers correct 
\end{itemize}

## The Randomization Distribution of $X$

\begin{center}
\only<1>{\includegraphics[angle=0, width=4in]{figs/02-riHistogram}}
\only<2>{\includegraphics[angle=0, width=4in]{figs/02-riHistogram2}}
\only<3>{\includegraphics[angle=0, width=4in]{figs/02-riHistogram3}}
\end{center}


## Parametric Null Hypothesis Significance Testing

\Large

- Specify and assume $H_0$
- Define $H_A$
- Examine reference dist'n ($t$, $\chi^2$, \ldots) under $H_0$
- Calculate $p$-value
- Compare to some $\alpha$; reject $H_0$ if $p<\alpha$

# Overview of Randomization Inference

## Randomization Inference

\Large 

>- Specify and assume $H_0$  
$\quad$ (sharp null of no treatment effect)
>- Define $H_A$
>- Create reference dist'n from all possible values of $X$ under $H_0$  
$\quad$ (or at least a big sample of them)
>- What prop. of possible values are ``at least as extreme as\text{''} observed?  
$\quad \rightsquigarrow$ $p$-value!
>- Compare to some $\alpha$; reject $H_0$ if $p<\alpha$
>- CA ballot ordering effects (JASA 2006)


## Randomization Inference

\large 

\center
The RI $p$-value is $$p = \frac{\text{\# outcomes } \geq \text{as extreme as obs}}{\text{total \# outcomes}}$$

\pause 

or

$$p = \frac{\text{\# randomizations producing extreme } \widehat{ATE}}{\text{total \# randomizations}}$$

\pause 

\vspace{2mm}

How many randomizations are there?

# Counting Principles

## Combinations: Counting selected sets

How many ways to **select** $k$ things from a set of $n$ things?

$$_n C_k = \binom{n}{k} = \frac{_n P_k}{k!} = \frac{n!}{k!(n-k)!}$$
\pause

Suppose 5 units, $A$, $B$, $C$, $D$, $E$.

>- How many ways to order? $_n P_k$: $ABCDE$, $ABCED$, \ldots $= 5 \cdot 4 \cdot 3 \cdot 2 \cdot 1 = 5! = 120$
>- If only want 3 of 5? Divide by $2! = (n-k)!$  
(removing permutations from last 2 slots)
>- What if order doesn't matter? Divide by $3! = k!$  
(6 permutations for $ABC$, but only one combination)



## Combinations: Counting selected sets

How many ways to choose 5 villages of 10 for treatment? \pause 

$$_{10} C_5 = \binom{10}{5} = \frac{10!}{5!(10-5)!}$$
\pause 

$$\frac{10\cdot 9 \cdot 8 \cdot 7 \cdot 6}{5\cdot 4 \cdot 3 \cdot 2 \cdot 1} = 252$$


## Common Assumptions, Null Hypotheses

\large

\begin{itemize}
\item Constant effect:

$$\tau_i = Y_{i1}-Y_{i0} = \tau \quad \forall i$$

\item Null hypothesis of no average effect:

$$ATE = \overline{\tau} = 0$$

\item Sharp null hypothesis of no effect: 

$$\tau_i = 0$$
\end{itemize}

# Examples

## An Assignment Mechanism: Perfect Doctor

\large 

Calculate RI $p$-value for Perfect Doctor, under sharp null. 


| Patient | Y(0)        | Y(1)         | $\tau$       | T |
| :-----: | :---------: | :----------: | :----------: | - |
| 1       | \alert{(1)} | 6            | \alert{(5)}  | 1 |
| 2       | \alert{(3)} | 12           | \alert{(9)}  | 1 |
| 3       | 9           | \alert{(8)}  | \alert{(-1)} | 0 |
| 4       | 11          | \alert{(10)} | \alert{(-1)} | 0 |
||||||
| Mean    | 10          | 9            | \alert{(3)}  |   |


\pause 
\vspace{3mm}

(See `01-ri-perfect-dr.R`)

## RI versus the $t$-test

\Large

Perfect Doctor:

- RI: $p = 1$
- `t.test()`: $p \approx 0.8$
- "If no tr effect, then this result typical"

\pause

\vspace{4mm}

\large

(Odd logic of NHST: "assume false thing, how strange is data?")


## Randomization Inference, Example 2

Assumed table of potential outcomes:

\begin{center}
\begin{tabular}{ccccc}
	Village & $T$ & \% if Female & \% if Male &  $\tau_i$ \\
	& &  Head, $Y(1)$ &  Head, $Y(0)$ &  \\ \hline
	1 & $\_$ & 15 & 10 & 5\\
	2 & $\_$ & 15 & 15 & 0 \\
	3 & $\_$ & 30 & 20 & 10 \\
	4 & $\_$ & 15 & 20 & -5\\
	5 & $\_$ & 20 & 10 & 10 \\
	6 & $\_$ & 15 & 15 & 0 \\
	7 & $\_$ & 30 & 15 & 15\\ \hline
Average	 && 20 & 15 & 5
\end{tabular}
\end{center}


## Randomization Inference, Example 2

Suppose we randomly select 2 villages to have female-headed councils, and observe

\begin{center}
\begin{tabular}{ccccc}
	Village & $T$ & \% if Female & \% if Male &  $\tau_i$ \\
	& &  Head, $Y(1)$ &  Head, $Y(0)$ &  \\ \hline
	1 & F & 15 & & \\
	2 & M &  & 15 &  \\
	3 & M &  & 20 &  \\
	4 & M &  & 20 & \\
	5 & M &  & 10 &  \\
	6 & M &  & 15 & \\
	7 & F & 30 &  & \\ \hline
Average	 && 22.5 & 16 & 6.5
\end{tabular}
\end{center}


## Randomization Inference, Example 2

We assume the _sharp null_ hypothesis (assumed values in \textcolor{red}{red}): 

\begin{center}
\begin{tabular}{ccccc}
	Village & $T$ & \% if Female & \% if Male &  $\tau_i$ \\
	& &  Head, $Y(1)$ &  Head, $Y(0)$ &  \\ \hline
	1 & F & 15 & {\color{red} 15} & {\color{red} 0}\\
	2 & M & {\color{red} 15} & 15 & {\color{red} 0} \\
	3 & M & {\color{red} 20} & 20 & {\color{red} 0} \\
	4 & M & {\color{red} 20} & 20 & {\color{red} 0}\\
	5 & M & {\color{red} 10} & 10 & {\color{red} 0} \\
	6 & M & {\color{red} 15} & 15 &{\color{red} 0} \\
	7 & F & 30 & {\color{red} 30} & {\color{red} 0} \\ \hline
Average	 &&  &  & {\color{red} 0}
\end{tabular}
\end{center}


## Randomization Inference, Example 2

Then we estimate what the observed ATE would be for all the possible random assignments.

First,

\begin{center}
\begin{tabular}{ccccc}
	Village & $T$ & \% if Female & \% if Male &  $\tau_i$ \\
	& &  Head, $Y(1)$ &  Head, $Y(0)$ &  \\ \hline
	1 & F & 15 & & \\
	2 & M &  & 15 &  \\
	3 & M &  & 20 &  \\
	4 & M &  & 20 & \\
	5 & M &  & 10 &  \\
	6 & F & {\color{red} 15} &  & \\
	7 & M &  & {\color{red} 30} &  \\ \hline
Average	 && 15 & 19 & -4
\end{tabular}
\end{center}


## Randomization Inference, Example 2

Second,

\begin{center}
\begin{tabular}{ccccc}
	Village & $T$ & \% if Female & \% if Male &  $\tau_i$ \\
	& &  Head, $Y(1)$ &  Head, $Y(0)$ &  \\ \hline
	1 & F & 15 &  & \\
	2 & M &  & 15 &  \\
	3 & M & & 20 &  \\
	4 & M &  & 20 & \\
	5 & F & {\color{red} 10} &  &  \\
	6 & M &  & 15 & \\
	7 & M & & {\color{red} 30} &  \\ \hline
Average	 && 12.5 & 20 & -7.5
\end{tabular}
\end{center}


## Randomization Inference, Example 2

\ldots, \pause and all the others.  The full set of $\frac{7!}{2!5!} = 21$ differences in means:

\begin{center}
\begin{tabular}{ccc}
& Estimate & Frequency  \\ \hline
& -7.5 & 3 	   \\
& -4 & 5 \\
& -0.5 & 6 \\
& 3 & 2 \\
& 6.5 & 3 \\
& 10 & 2 \\ \hline
Total & & 21
\end{tabular}
\end{center}

\pause

How many are at least as extreme as my 6.5? \pause

\vspace{2mm}
Two-sided (``women's \% $\neq$ men's\text{''}): $p = \frac{8}{21} \approx 0.38$ \pause

\vspace{2mm}
One-sided (``women's \% $>$ men's\text{''}): $p = \frac{5}{21} \approx 0.24$


## Randomization Inference

\Large

These $p$-values

>- are exact, not approximate
>- do not require asymptotic large $N \to \infty$
>- need no distrib assump (normal as $N \to \infty$)
>- do not require calculating standard error (SE)
>- _Randomization_ creates distribution of possible numbers correct/treatment effects
>- can be approximated by sampling
>    - Assigning 500 of 1000 respondents to treatment?  Randomly generate 10,000 assignments from $2.7 \times 10^{299}$ possible \ldots
>- can be computationally intensive


## Randomization Inference

\large

- Resume audit study, @bermul04

```{r}
#| label: readresume
#| echo: false
#| message: false
#| warning: false

data(resume, package = "qss")
kable(table(resume$race, resume$call))
```

\pause 

- Only possible values: $\tau_i \in \{-1, 0, 1\}$

\pause 

```{r}
#| message: false
#| warning: false

resume |> group_by(race) |>
  summarise(call_rate = mean(call))
```


***

\large

>- Assume the sharp null $\tau_i = 0$ for every employer.
>- $H_0: \mu_{\text{black name}} = \mu_{\text{white name}}$
>- $H_A: \mu_{\text{black name}} \neq \mu_{\text{white name}}$
>- Create reference dist'n of all possible assignments

\pause 

$$_{4870} C_{2435} = \binom{4870}{2435} = \frac{4870\cdot 4869 \cdot \ldots \cdot 2436}{2435!}$$ 

\pause 

$\approx 1.1 \times 10^{1464}$

\pause 

(There are $\approx 10^{86}$ fundamental particles in the universe.)

\pause 

>- Let's do 1000, or 100,000 -- something reasonable
>- See `01-ri-resume-donate.R`


***

\begin{center}
\only<1>{\includegraphics[angle=0, width=3.5in]{figs/02-riHist_resume1}}
\only<2>{\includegraphics[angle=0, width=3.5in]{figs/02-riHist_resume2}}
\end{center}

## Randomization Inference

\Large

- @gergre12 donations example, p. 65
- Possible values $\tau_i \in (-\infty, \infty)$
- $Y_1$, $Y_0$, $\tau$ likely very skewed
- See `01-ri-resume-donate.R`


***

\begin{center}
\only<1>{\includegraphics[angle=0, width=3.5in]{figs/02-riHist_donate1}}
\only<2>{\includegraphics[angle=0, width=3.5in]{figs/02-riHist_donate2}}
\end{center}

## The RI Confidence Interval

\large

Recall that 

\begin{center}
``reject $H_0$ at $\alpha = 0.05$'' $\equiv$ ``$H_0$ falls outside 95\% CI''
\end{center}

\pause 
\vspace{5mm}

Create RI confidence ints ("invert the randomization test"):

>- Posit $H_0: \tau = \tau^* \in \{ \ldots, -2, -1, 0, 1, 2, \ldots \}$
>- RI test whether to reject $H_0$
>- If not, then $\tau^*$ is in CI
>- CI consists of set of $\tau^*$ not unusual, given data
>- See `01-ri-resume-donate.R`

***

\LARGE

We can adapt this testing procedure for any quantity we want to test!

\pause 

>- Difference in variances
>- Regression coefficient
>- Sum of two regression coefficients
>- Ratio of skewnesses
>- \ldots (any crazy thing!)

\pause 
No need to derive the correct asymptotic standard error (SE).


***

\Huge

\center
"There is only one test."

\pause 
\vspace{1cm}

\huge 


-- Allen Downey, posit::conf(2024)

***

\huge

\begin{center}
Next:\\
Covariates in Experiments
\end{center}

***

