---
title: | 
    | Randomized Experiments with Clusters
author: "Ryan T. Moore"
date: 2024-08-21
date-format: iso
execute: 
  echo: true
format: 
  beamer:
    fonttheme: serif
    include-in-header:
      - text: | 
          \usepackage{wasysym} 
      - file: zzz_beamer_header.tex
    section-titles: true
    toc: true
institute:
  - American University
bibliography: "../admin/main.bib"
---

```{r}
#| label: setup
#| echo: false
#| message: false
#| results: false
#| warning: false

library(here)
library(tidyverse)
```

# Clusters in Experiments

## Clusters in Experiments

\large

>- So far, _individual units_ assigned to treatment
>- But, sometimes, _groups_ of units assigned to treatment. 
>- Consider assigning \ldots
>    - Classrooms (not students) to new curriculum
>    - Schools (not teachers or students) to new curriculum
>    - Media markets (not viewers) to ad campaign
>    - Clinic catchment areas (not patients) to health infrastructure spending
>- These groups of individuals are _clusters_
>- (Let's start with easy case: all clusters same size \ldots)  
[@gergre12]

# Variance

## Variance

\large 

- CLT SE for diff in means for clustered design:

\pause 

\scriptsize

$$SE(\widehat{ATE}) = \sqrt{\frac{1}{k-1} \left[ \frac{m \mathrm{Var}(\overline{Y_j}(0))}{N-m} + \frac{(N-m)\mathrm{Var}(\overline{Y_j}(1))}{m} +
2\mathrm{Cov}(\overline{Y_j}(0), \overline{Y_j}(1)) \right]}$$

\pause 

\normalsize 

What might make this SE large?

>- Dividing by $k \equiv$ number of _clusters_
>- So, small number of clusters $\rightsquigarrow$ larger SE
>- Small clusters (more variability in means)
>- If clusters differ in potential outcomes $\rightsquigarrow$ larger SE
>- I.e., if clusters meaningful $\rightsquigarrow$ larger SE

## Variance

\large 

- CLT SE for diff in means for clustered design:

\scriptsize

$$SE(\widehat{ATE}) = \sqrt{\frac{1}{k-1} \left[ \frac{m \mathrm{Var}(\overline{Y_j}(0))}{N-m} + \frac{(N-m)\mathrm{Var}(\overline{Y_j}(1))}{m} +
2\mathrm{Cov}(\overline{Y_j}(0), \overline{Y_j}(1)) \right]}$$

\normalsize

What might make this SE small?

>- Many clusters
>- If clusters look like each other, $\rightsquigarrow$ smaller SE
>- $\mathrm{Var}(\overline{Y_j}(1))$: variance in cluster averages of $Y_1$
>- $\mathrm{Var}(\overline{Y_j}(0))$: variance in cluster averages of $Y_0$
>- If $\mathrm{Var}(\overline{Y_j}(1)), \mathrm{Var}(\overline{Y_j}(0))$ are small, SE is small

# Bias

## Bias

\large

>- Consider case where clusters vary in size, and size correlates with $Y_1$, $Y_0$
>    - Classrooms differ in number of students, and larger classes have lower test scores
>    - Media markets vary in number of viewers, and voters in cities less likely to turn out
>    - Health clusters vary in patient population, and larger clusters are healthier
>- In this case, individual difference in means estimator is _biased_ for true ATE
>    - (Different assignments of clusters will produce different counts of treated units!)


## Bias

\large

Alternative estimator: Difference in totals (not means)

$$\widehat{ATE} = \frac{k_T + k_C}{N} \left(\frac{\sum Y_i(1) | T_i = 1}{k_T} - \frac{\sum Y_i(0) | T_i = 0}{k_C} \right)$$

\pause 

>- But, doesn't include cluster sizes, so may be high variance
>- I.e., more students per classroom doesn't help precision here (unlike above)

# An Example

## An Example

\large

>- 1000 voters, split across 10 cities
>- Measure ideology, which varies by city and voter
>- Do 2 hypothetical assignments
>    - Individual-level assignment to news
>    - City-level cluster assignment to news
>- Use RI to find SE in each case

## An Example

```{r}
library(fabricatr)
library(randomizr)
library(tidyverse)

# Make data and visualise

set.seed(95852894)

voters <- fabricate(
  N = 1000,
  city_id = rep(1:10, 100),
  ideology = draw_normal_icc(mean = 0, N = N, 
                             clusters = city_id, ICC = 0.7),
  city_id_fac = as.factor(city_id),
  city_id_fac = fct_reorder(city_id_fac, ideology)
  )
```

## An Example

```{r}
table(voters$city_id)
```

\pause 

\scriptsize
```{r}
voters |> group_by(city_id) |>
  summarise(mean_ideo = mean(ideology)) |>
  arrange(mean_ideo)
```


## An Example

```{r}
#| echo: false
voters |> ggplot(aes(city_id_fac, ideology)) +
  geom_violin(draw_quantiles = 1/2) +
  labs(x = "City", y = "Ideology")
```

## An Example

Assignments

```{r}
voters <- voters |> 
  mutate(
    tr_ind = sample(rep(0:1, nrow(voters) / 2)),
    tr_cl = cluster_ra(clusters = voters$city_id, m = 5))
```

\pause 

```{r}
voters |> count(tr_ind)
voters |> count(tr_cl)
```

## An Example

\large
Conditions by city, individual-level assignment:

```{r}
table(voters$city_id, voters$tr_ind)
```

## An Example

\large
Conditions by city, cluster-level assignment:

```{r}
table(voters$city_id, voters$tr_cl)
```

## An Example

\large
Draw responses under each assignment:

```{r}
voters <- voters |>
  mutate(
    response_ind = draw_likert(x = ideology + tr_ind, min = -7, max = 7, bins = 7), 
    response_cl = draw_likert(x = ideology + tr_cl, min = -7, max = 7, bins = 7))
```

\pause 

```{r}
#| echo: false
#| fig-height: 5

voters |> ggplot(aes(ideology, jitter(response_cl))) +
  geom_point(aes(colour = city_id_fac,
                 shape = factor(tr_cl))) +
  labs(x = "Latent Ideology Score",
       y = "Likert Survey Response (jittered)")
```

## An Example

\large

Analyse precision with RI:

>- Do 1000 individual-level hypothetical assignments
>- Do 1000 cluster-level hypothetical assignments
>- Use sharp null $H_0: \tau_i = 0$ to impute unobserved potential outcomes
>- Calculate $\widehat{ATE}$ in each case
>- Compare distribution of $\widehat{ATE}_{\text{Individual}}$ to  
distribution of $\widehat{ATE}_{\text{Cluster}}$
>- With partner, sketch the two randomization distributions!

## An Example

Analyse precision with RI:

\footnotesize

```{r}
n_sims <- 1000
df_ates <- tibble(ate_ind = NA,
                  ate_cl = NA)

for(idx in 1:n_sims){
  
  voters <- voters |> 
    mutate(
      hyp_tr_individ = sample(rep(0:1, nrow(voters) / 2)),
      hyp_tr_cl = cluster_ra(clusters = voters$city_id, m = 5))
  
  ate_ind <- mean(voters$response_ind[voters$hyp_tr_ind == 1]) -
    mean(voters$response_ind[voters$hyp_tr_ind == 0]) 
  
  ate_cl <- mean(voters$response_cl[voters$hyp_tr_cl == 1]) -
    mean(voters$response_cl[voters$hyp_tr_cl == 0]) 
  
  df_ates[idx, "ate_ind"] <- ate_ind
  df_ates[idx, "ate_cl"] <- ate_cl
}
```

## An Example

Analyse precision with RI:

\pause 

```{r}
#| echo: false
#| fig-height: 5

df_long <- df_ates |> pivot_longer(everything()) |>
  mutate(Assignment = if_else(name == "ate_ind", "Individual", "Cluster"))

ggplot(df_long, aes(value)) + 
  geom_density(aes(color = Assignment,
                   linetype = Assignment),
               linewidth = 2) +
  labs(x = "Hypothetical Estimated ATE")
```

\pause 

```{r}
#| echo: false

df_ates |> summarise(se_ind = sd(ate_ind),
                     se_cl = sd(ate_cl))
```




<!-- Include a PDF/PNG/... -->
<!-- ![](figs/myfig.pdf){fig-align="center" height=80%} -->

<!-- Include PDFs in columns -->
<!-- \includegraphics[width=2in]{figs/myfig1.pdf} -->
<!-- \includegraphics[width=2in]{figs/myfig2.pdf} -->

***

\huge

\begin{center}
Next:\\
Regression and Experiments
\end{center}

\vspace{5mm}

\large

\center
`rtm@american.edu`  
`www.ryantmoore.org`  

\quad

<!-- \vspace{5mm} -->
<!-- _Journal of Public Policy_   -->
<!-- \hspace{3.5cm} @mooganmin22 -->

## References {.allowframebreaks}

\footnotesize

