---
title: | 
    | Interference 
header-includes:
  - \usepackage{datetime}
  - \usepackage{amsmath}
  - \usepackage{wasysym}
  - \usepackage{color}
date: "2024-08-23"
author: Ryan T. Moore
institute: American University
output:
  beamer_presentation:
    slide_level: 2
    toc: true
    fonttheme: serif
    includes:
      in_header: zzz_beamer_header.tex
  pdf_document:
    fig_caption: true
    number_sections: true
    urlcolor: blue
link-citations: yes 
bibliography: "../admin/main.bib"
---
  
```{r knitr options, echo=FALSE, warning=FALSE}
# Encoding required for proper knitr rendering
options(encoding = "native.enc")
```

```{r enviro, warning = FALSE, results = 'hide', echo = FALSE, message = FALSE}

library(blockTools)
library(xtable)
library(knitr)
library(png)
library(ggdag)
library(grid)
library(gridExtra)
library(devtools)
library(tidyverse)
```

# Interference

## Interference


\Large

@rubin78 
 
SUTVA 2: No interference between units.

Does $A$'s assignment affect $B$'s potential outcomes? 


## Interference

\large 
>- Describe violation of SUTVA in social pressure experiment.  
>- Does assg of MTurk survey respondent $A$ to conjoint profile $X$ affect response of MTurker $B$? 
>- Does assg of election monitors to polling place $A$ affect fraud at
    $B$?
>- Does assg of child $A$ to vaccine affect health of $B$?  
>- Does assg of $A$ to GOTV door knock affect turnout of neighbor?
>- Does assg of develpmnt £ to village $A$ affect trust in $B$? 
>- Other common examples?

## Examples of Interference in Field Experiments

@gergre12

\large

>- Contagion
>- Displacement
>- Communication
>- Knowledge of others (disappointment/pride)
>- Deterrence 
>- Persistence 
>- Memory


## Bias under Interference

| Agent | Y if MaryH | Y if PeterH | Y if LimorH | Y if NobodyH |
| :---------: | :--------: | :---------: | :---------: | :----------: |
|    Mary     |    100     |     50      |     70      |      70      |
|    Peter    |     50     |     50      |     50      |      50      |
|    Limor    |     90     |     50      |     90      |      90      |
| --------- | -------- | --------- | --------- | ---------- |
|    Mean    |     80     |     50      |     70      |      70      |

\pause 
>- $Y_i(0)$ _unstable_, since depends on who assigned Tr
>- $\overline{Y}_{\textrm{None}} = 70$.
>- Calc TEs relative to $\overline{Y}_{\textrm{None}}$ ("uniformity")

## Bias under Interference: Uniformity Trial

| Agent | Y if MaryH | Y if PeterH | Y if LimorH | Y if NobodyH |
| :---------: | :--------: | :---------: | :---------: | :----------: |
|    Mary     |    100     |     50      |     70      |      70      |
|    Peter    |     50     |     50      |     50      |      50      |
|    Limor    |     90     |     50      |     90      |      90      |
| --------- | -------- | --------- | --------- | ---------- |
|    Mean    |     80     |     50      |     70      |      70      |


>- True uniformity ATE: $\frac{(100-70) + (50 - 50) + (90-90)}{3} = 10$ \pause 
>- But, calculate all TEs relative to uniformity: 
>    - If Mary, $\frac{100+50+90}{3} - 70 = 80-70 = 10$
>    - If Peter, $\frac{50+50+50}{3} - 70 = 50 - 70 = -20$
>    - If Limor, $\frac{70+50+90}{3} - 70 = 70-70 = 0$
>- So, $E(\textrm{Diff in Means}) = \frac{10-20+0}{3} = -\frac{10}{3}$
>- So, $E(\textrm{Diff in Means}) \neq ATE$
>- Diff in means is _biased_ for uniformity ATE! \pause  \frownie



## Bias under Interference: Difference-in-Means

| Agent | Y if MaryH | Y if PeterH | Y if LimorH | Y if NobodyH |
| :---------: | :--------: | :---------: | :---------: | :----------: |
|    Mary     |    100     |     50      |     70      |      70      |
|    Peter    |     50     |     50      |     50      |      50      |
|    Limor    |     90     |     50      |     90      |      90      |
| --------- | -------- | --------- | --------- | ---------- |
|    Mean    |     80     |     50      |     70      |      70      |

>- What about empirical difference in means, $\widehat{\textrm{Diff in Means}}$?
>- Empirically, if randomly select winner, calc diff-in-means,
>    - If Mary, $100-\frac{50+90}{2} = 100-70 = 30$
>    - If Peter, $50 - \frac{50+50}{2} = 50 - 50 = 0$
>    - If Limor, $90-\frac{70+50}{2} = 90-60 = 30$
>- $E(\widehat{\textrm{Diff in Means}}) = 20$

\pause 

$E(\widehat{ATE}) \neq ATE$, so, estimator _biased_ under randomization. \pause \frownie

## Bias under Interference: Reference Control

| Agent | Y if MaryH | Y if PeterH | Y if LimorH | Y if NobodyH |
| :---------: | :--------: | :---------: | :---------: | :----------: |
|    Mary     |    100     |     50      |     70      |      70      |
|    Peter    |     50     |     50      |     50      |      50      |
|    Limor    |     90     |     50      |     90      |      90      |
| --------- | -------- | --------- | --------- | ---------- |
|    Mean    |     80     |     50      |     70      |      70      |

>- What if use "Limor honored" as reference "control"?
>- True "Limor" ATE: $\frac{(100-70) + (50 - 50)}{2} = 15$
>- Calculate diff-in-means TE's relative to "Limor honored"
>    - If Mary, $80-70=10$
>    - If Peter, $50-70=-20$
>    - $E(\textrm{Diff in Means}) = \frac{10-20}{2} = -5$
>- And, $E(\widehat{\textrm{Diff in Means}}) = 20$

\pause 

With interference, diff-in-means under randomization biased. \pause \frownie


## Notation

\large

\begin{eqnarray*}
Y_i(z_i, {\bf z}) & \stackrel{?}{=} & Y_i(z_i, {\bf z}') \\ \pause
Y_1\left(1, \left(1, 0, 0, 1, \ldots \right) \right) &
\stackrel{?}{=} & Y_1\left(1, \left(1, 1, 0, 0, \ldots \right) \right)\\
Y_0\left(0, \left(0, 0, 0, 1, \ldots \right) \right) &
\stackrel{?}{=} & Y_0\left(0, \left(0, 1, 0, 0, \ldots \right) \right)
\end{eqnarray*}




## Estimation in Presence of Interference

>- Let $T_i'=$ whether my friend is treated
>- $T_i=$ whether I (unit $i$) am treated 
>- Then, $Y_{T_i'T_i}$ is my potential outcome using both assignments  

\pause 

Under balanced random assignment, **no** interference, naive diff in means has expectation 

\pause 

$$\frac{\bar{Y}_{11} + \bar{Y}_{01}}{2} - \frac{\bar{Y}_{10} + \bar{Y}_{00}}{2} = \bar{Y}_{01} - \bar{Y}_{00}$$  

\pause 

which is _unbiased_ for ATE.


## Estimation in Presence of Interference


\large


>- But, if $$\underbrace{\bar{Y}_{11} - \bar{Y}_{01}}_{\textrm{spillover eff on Tr}} > \underbrace{\bar{Y}_{10} - \bar{Y}_{00}}_{\textrm{spillover eff on Co}}$$ then tend to overest $\bar{Y}_{01} - \bar{Y}_{00}$
>- If $$\underbrace{\bar{Y}_{11} - \bar{Y}_{01}}_{\textrm{spillover eff on Tr}} < \underbrace{\bar{Y}_{10} - \bar{Y}_{00}}_{\textrm{spillover eff on Co}}$$ then tend to underestimate $\bar{Y}_{01} - \bar{Y}_{00}$.



## Design for Estimating Interference

\large

>- Multilevel experiments can help estimate these effects
>    - Given individuals within communities 
>    - Randomly assign whether community gets 20% or 80% treatment
>    - Randomly assign treatment to some individuals in communities 
>- ``Detecting Spillover Effects: Design and Analysis of Multilevel Experiments\text{''}, @sinmccgre12

## Design for Estimating Interference

\large

OES project with IRS (US tax authority): 

\pause 

>    - assign preparers to "some client letters", "no client letters"
>    - assign clients in "some client letters" preparers to letters/no letters



## Diagnosis of Potential Interference

1. Block:

```{r warning=FALSE}
data(x100)
block.out <- block(data = x100, groups = "g", 
                   id.vars = c("id"), 
                   block.vars = c("b1", "b2"))
```

2. Assign:

```{r warning = FALSE}
assg.out <- assignment(block.out, seed = 157)
```

## Diagnosis of Potential Interference

Diagnose interference after assgnmnt (1D, @lintsicoh01)

3. \alert{Diagnose}: 

```{r warning = FALSE}
diagnose(assg.out, data = x100, id.vars = "id", 
    suspect.var = "b1", suspect.range = c(0, 5))
```

## Further Examination of Design

4. Get block IDs:

```{r warning = FALSE}
createBlockIDs(assg.out, data = x100, id.var = "id")
```

## Further Examination of Design

5. Get balance:

```{r warning = FALSE}
assg2xBalance(assg.out, x100, id.var = "id", 
    bal.vars = c("b1", "b2"))
```


## Avoiding Potential Interference due to Proximity

Are units too near each other?

\pause 

\includegraphics[width=4.8in]{figs/04-rand3k.pdf}



<!-- ## Avoiding Potential Interference due to Proximity -->

<!-- \includegraphics[width=4.8in]{figs/14-rand3k2.pdf} -->


## Avoiding Potential Interference due to Proximity

3000 iterations, max min distance:

\includegraphics[width=4.8in]{figs/04-maxmin3k.pdf}



## Estimating Direct and Indirect Effects

\large

>- Different to mediation _direct_, _indirect_ effects

```{r warning = FALSE, echo = FALSE, fig.height=4.3}
coords <- list(x = c(T = 0, Y = 2, M = 1), 
               y = c(T = 1, Y = 1.1, M = 1.5))
dagify(Y ~ T + M, M ~ T, coords = coords) %>% 
  ggdag_classic(size = 30) + theme_dag_blank()
```

## Estimating Direct and Indirect Effects

\large

@hudhal08 (on @aliemcvon05)

>- Here, 
>    - _direct_: from treating $i$ 
>    - _indirect_: from treating $j \neq i$

\pause 

Think of this as different problem 

>- (Though, "effect of treatment through others" works?)
>- (Common concept, but not really a mediating _variable_)

## Estimating Direct and Indirect Effects

@hudhal08 (on @aliemcvon05)

\vspace{-20mm}

```{r echo=FALSE}
img <- readPNG("figs/04-hudhal-table-1.png")
grid.raster(img)
```

\vspace{-15mm}

>- Consider $>50$\% as $A$, $<28$\% as $B$  
>- Direct effect in $B$: $7.01-2.66 = 4.35$  
>- Direct effect in $A$: $1.47-1.27=0.2$  
>- Indirect effect on unvaccinated: $7.01-1.47= 5.54$  
>- Total effect: $7.01-1.27=5.74$ 
>- Overall effect: $35/8479-25/18623 = 2.79/1000$

## Estimating Direct and Indirect Effects

\large 

@hudhal08 (on @aliemcvon05)

\Large
	
Designing the randomized experiment:

```{r echo=FALSE}
img <- readPNG("figs/04-hudhal-table-2.png")
grid.raster(img)
```

## @hudhal08 (on Ali 2005)

```{r echo=FALSE}
img <- readPNG("figs/04-hudhal-table-3.png")
grid.raster(img)
```

\vspace{-10mm}

- $\psi$: 50% coverage
- $\phi$: 30% coverage


## Application: Displacing Election Irregularities

\large 

@ichsch12 in Ghana

Design:

>- Blocks of 3 constituencies, select 1 for Tr, 2 for Co
>- Treatment constituencies: send observers to 25% of areas
>- So, areas can be 
>    - Tr in Tr constituency
>    - Co in Tr constituency
>    - Co in Co constituency

>- Where election observers are sent, smaller registration irregularities
>- In nearby control areas, _larger_ irregularities

## @sobel06

\large 

If interference, diff-in-means estimator (or regression coef) 

- is **not** unbiased for ATE
- is difference:

$$(\textrm{ITT for Tr group}) - (\textrm{indirect/spillover effect on Co group})$$


## @rosenbaum07

>- Randomization inference gives valid coverage, even if interference

Null Hypotheses:

>- No primary effect:  $H_0: Y_{bi\mathbf{z}} = Y_{bi\mathbf{z}'}$
>    - no effect of treatment itself on $i$
>- No effect: $H_0: Y_{bi\mathbf{z}} = \widetilde{Y}_{bi}$
>    - Outcome for $i$ is same as if assigned Tr, but Tr withheld from _everyone_ ("uniformity")

\pause 

Effect, but no Primary Effect: 

>- if exposing block gets everyone sick, then no primary effect of $i$ getting directly exposed
>- if news raises anxiety in HH, irrelevant if I saw news



***

\huge

\begin{center}
Next: Causal Forests
\end{center}

***


## References {.allowframebreaks}

\footnotesize




